---
title: "Use cases for <i>Interactive Causal Structure Discovery in Earth System Sciences</i>"
output: html_notebook
---

```{r, include=FALSE}
#' THIS SOURCE CODE IS SUPPLIED “AS IS” WITHOUT WARRANTY OF ANY KIND, AND ITS
#' AUTHOR AND THE JOURNAL OF MACHINE LEARNING RESEARCH (JMLR) AND JMLR’S
#' PUBLISHERS AND DISTRIBUTORS, DISCLAIM ANY AND ALL WARRANTIES, INCLUDING BUT
#' NOT LIMITED TO ANY IMPLIED WARRANTIES OF MERCHANTABILITY AND FITNESS FOR A
#' PARTICULAR PURPOSE, AND ANY WARRANTIES OR NON INFRINGEMENT. THE USER ASSUMES
#' ALL LIABILITY AND RESPONSIBILITY FOR USE OF THIS SOURCE CODE, AND NEITHER THE
#' AUTHOR NOR JMLR, NOR JMLR’S PUBLISHERS AND DISTRIBUTORS, WILL BE LIABLE FOR
#' DAMAGES OF ANY KIND RESULTING FROM ITS USE. Without limiting the generality
#' of the foregoing, neither the author, nor JMLR, nor JMLR’s publishers and
#' distributors, warrant that the Source Code will be error-free, will operate
#' without interruption, or will meet the needs of the user.

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = T)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)

set.seed(125)

setwd('../code/')
source('global.R')
```

## Helper functions for plotting results
```{r, fig.width=12, fig.height=3}
palette <- brewer.pal(3, 'Set2')


plot.difference.graph <- function (adj.m, compare.adj.m, tag) {
  par(mar=rep(0,4), cex=2)

  graph.a <- graph_from_adjacency_matrix(adj.m)
  varnames <- colnames(adj.m)
  missing.edges <- as.vector(apply(which(adj.m == 0 & t(adj.m) == 0 & compare.adj.m == 1, arr.ind=T), 1, function (idx) c(varnames[idx[1]], varnames[idx[2]])))
  graph.a <- add_edges(graph.a, missing.edges)

  colors <- c()
  if (sum(adj.m) > 0)
    colors <- apply(which(t(adj.m) == 1, arr.ind=T), 1, function (idx) {
      if (t(compare.adj.m)[idx[1], idx[2]] == 1) 'black'
      else if (t(compare.adj.m)[idx[2], idx[1]] == 1) palette[3]
      else palette[2]
    })

  plot(
    graph.a, layout=layout_in_circle(graph.a), vertex.size=70, vertex.color=NA, edge.width=5,
    edge.color=c(colors, rep(palette[1], length(missing.edges))), vertex.label.family='sans',
    vertex.label.color='black', vertex.width=3.5,
    edge.lty=c(rep(1, length(colors)), rep(2, length(missing.edges))), margin=rep(0.1, 4)
  )

  text(-1.2, 1.1, tag, cex=1.5)
}


plot.strip <- function (graphs, fn, legend=F) {
  rows <- ceiling(length(graphs) / 5)
  pdf(file=paste0(fn, '.pdf'), width=20, height=4 * rows)
  par(mfrow=c(rows, 5))

  invisible(sapply(seq_along(graphs), function (i) {
    if (i < length(graphs)) plot.difference.graph(graphs[[i]], graphs[[i + 1]], tag=paste0(i, ')'))
    else plot.difference.graph(graphs[[i]], graphs[[i]], tag=paste0(i, ')'))
  }))

  if (legend) {
    plot(NULL, xaxt='n', yaxt='n', bty='n', xlim=0:1, ylim=0:1, ylab='', xlab='')
    legend('center', legend=c('Reversal', 'Addition'), col=palette[c(3, 1)], lty=c(1,2), lwd=8, bty='n', seg.len=3)
  }

  dev.off()
}

plot.cv <- function (graphs, data, k, fn, last.fold=NA, legend=F) {
  if (!is.na(last.fold)) folds <- c(cut(seq(nrow(data) - last.fold), k - 1, labels=F), rep(k, last.fold))
  else folds <- cut(seq(nrow(data)), k, labels=F)

  titles <- lapply(graphs, function (g) {
    sapply(seq(k), function (i) {
      tr.idx <- which(folds != i)
      calc.mean.R2(data, tr.idx, g, train=T)
    })
  })

  plot.strip(graphs, fn, legend=legend)

  titles <- reshape2::melt(titles)
  colnames(titles) <- c('type', 'fold', 'value', 'step')
  titles
}

plot.r2 <- function (data, legend=F) {
  data$size <- 0.5
  agg.d <- aggregate(data$value, list(data$type, data$step), mean)
  data <- rbind(data,
                data.frame(type=agg.d$Group.1, fold=rep(0, nrow(agg.d)),
                           value=agg.d$x, step=agg.d$Group.2, size=rep(1, nrow(agg.d))))
  steps <- seq(1, max(data$step))

  p <- ggplot(data, aes(x=step, y=value, color=type)) +
    geom_line(aes(group=interaction(fold, type), size=size, alpha=size)) +
    labs(x='Step', y='Mean adjusted R-squared') +
    scale_color_manual(name='Data', values=c('red', 'black')) +
    scale_size_continuous(name='Type', breaks=c(0.5, 1), labels=c('Fold', 'Mean'), range=c(0.5, 1)) +
    scale_alpha_continuous(name='Type', breaks=c(0.5, 1), labels=c('Fold', 'Mean'), range=c(0.5, 1)) +
    scale_x_continuous(breaks=steps) +
    theme_minimal() +
    theme(legend.position='top', text = element_text(size=20))

  if (legend) {
    tmp <- ggplot_gtable(ggplot_build(p))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    ggsave('../figures/R2_legend.pdf', tmp$grobs[[leg]], width=10, height=1)
  }

  p + theme(legend.position = 'none')
}
```

## Helper functions for finding a path of single edits between two graphs
```{r}
find.model <- function (data, correct) {
  graphs <- list()
  current <<- initcurrent(data=data)

  graphs[[1]] <- current$edit.mtx

  rev.edges <- which(t(current$edit.mtx) == 1 & correct == 1, arr.ind=T)

  if (nrow(rev.edges) != 0)
    invisible(apply(rev.edges, 1, function (edge) {
      reverse.edge(edge[1], edge[2])
      graphs <<- append(graphs, list(current$edit.mtx))
    }))

  add.edges <- which(current$edit.mtx == 0 & correct == 1, arr.ind=T)

  if (nrow(add.edges) != 0)
    invisible(apply(add.edges, 1, function (edge) {
      add.edge(edge[2], edge[1])
      graphs <<- append(graphs, list(current$edit.mtx))
    }))

  del.edges <- which(current$edit.mtx == 1 & correct == 0, arr.ind=T)

  if (nrow(del.edges) != 0)
    invisible(apply(del.edges, 1, function (edge) {
      delete.edge(edge[1], edge[2])
      graphs <<- append(graphs, list(current$edit.mtx))
    }))

  graphs
}

find.plot.cv <- function (data, correct, k, fn, last.fold=NA, legend=F) {
  graphs <- find.model(data, correct)
  plot.cv(graphs, data, k, fn, last.fold=last.fold, legend=legend)
}

```

## Load data
```{r}
vars <- c('Rg', 'T', 'NEE', 'VPD', 'H', 'LE')
correct <- matrix(0, 6, 6, dimnames=list(vars, vars))
correct[cbind(c(1,1,1,1,2,2,4,4), c(2,3,5,6,3,4,3,6))] <- 1

hyy <- read.csv('../data/FI_HYY_Preprocessed_2014.csv')
hyy.apr <- hyy[hyy$MONTH == 4, vars]

hyy.long <- read.csv('../data/FI_HYY_Preprocessed_2013-2015.csv')
hyy.l.apr <- hyy.long[hyy.long$MONTH == 4, vars]
hyy.apr.may <- hyy.long[hyy.long$MONTH %in% c(4, 5), vars]
hyy.l.aug <- hyy.long[hyy.long$MONTH == 8, vars]

hyy.3apr.1aug <- rbind(hyy.l.apr, hyy.l.aug[(nrow(hyy.l.aug) - 214):nrow(hyy.l.aug), ])
```

# Use Case 1:

## Hyytiälä 1 year, April
```{r}
fn <- '../figures/expert_april14'
titles <- find.plot.cv(hyy.apr, correct, 2, fn, legend=T)

plot.r2(titles, legend=T) + ylim(NA, 1)
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

## Hyytiälä 3 years, April-May, 3 folds
```{r}
fn <- '../figures/expert_april-may_13-15_3'
titles <- find.plot.cv(hyy.apr.may, correct, 3, fn)

plot.r2(titles) + ylim(0, 1)
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

# Use Case 2:
## Hyytiälä 3 years April, 1 year August, 4 folds
```{r}
fn <- '../figures/expert_april13-15_aug15'
titles <- find.plot.cv(hyy.3apr.1aug, correct, 4, fn, last.fold=215)

plot.r2(titles) + ylim(NA, 1) + annotate('text', 9.75, -4.4, label='August 2015')
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

## Hyytiälä 3 years, April 3 folds
```{r}
fn <- '../figures/expert_april13-15'
titles <- find.plot.cv(hyy.l.apr, correct, 3, fn)

plot.r2(titles) + ylim(-.01, 1)
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

# Use Case 3:

## Function to compute the effect of all possible edits on the model
```{r}
current.state <- function (data) {
  all.improv <- list(Add=matrix(0, length(vars), length(vars), dimnames=list(vars, vars)),
                     Del=matrix(0, length(vars), length(vars), dimnames=list(vars, vars)),
                     Rev=matrix(0, length(vars), length(vars), dimnames=list(vars, vars)))
  curr.score <- 0

  sapply(seq(k), function (i) {
    tr.idx <- which(folds != i)
    i <- calc.improvements(data, tr.idx, current$edit.mtx, current$calc.score, all=T)
    all.improv$Add <<- all.improv$Add + i$Add
    all.improv$Del <<- all.improv$Del + i$Del
    all.improv$Rev <<- all.improv$Rev + i$Rev
    curr.score <<- curr.score + current$calc.score(data, tr.idx, current$edit.mtx)
  })

  all.improv$Add <- round(all.improv$Add / k, 5)
  all.improv$Del <- round(all.improv$Del / k, 5)
  all.improv$Rev <- round(all.improv$Rev / k, 5)
  curr.score <- round(curr.score / k, 5)

  print(curr.score)
  all.improv
}
```

## April-May 2013-2015, no knowledge
```{r}
set.seed(123)
data <- hyy.apr.may
k <- 3
folds <- cut(seq(nrow(data)), k, labels=F)

graphs <- list()
current <<- initcurrent(data=data)

graphs[[1]] <- current$edit.mtx
add.edge(3, 1)
graphs[[2]] <- current$edit.mtx
add.edge(2, 1)
graphs[[3]] <- current$edit.mtx
delete.edge(1, 4)
graphs[[4]] <- current$edit.mtx
add.edge(2, 5)
graphs[[5]] <- current$edit.mtx
reverse.edge(4, 2)
graphs[[6]] <- current$edit.mtx
add.edge(4, 1)
graphs[[7]] <- current$edit.mtx
delete.edge(5, 4)
graphs[[8]] <- current$edit.mtx
delete.edge(1, 2)
graphs[[9]] <- current$edit.mtx
reverse.edge(1, 5)
graphs[[10]] <- current$edit.mtx

fn <- '../figures/non_expert_alg'
titles <- plot.cv(graphs, data, k, fn)

plot.r2(titles) + ylim(0, 1)
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

## April-May 2013-2015, no knowledge, empty graph
```{r}
set.seed(123)
graphs <- list()
current <<- initcurrent(data=data)
current$edit.mtx <- matrix(0, ncol(data), ncol(data), dimnames=list(vars, vars))

graphs[[1]] <- current$edit.mtx
add.edge(5, 1)
graphs[[2]] <- current$edit.mtx
add.edge(2, 4)
graphs[[3]] <- current$edit.mtx
add.edge(6, 3)
graphs[[4]] <- current$edit.mtx
add.edge(4, 1)
graphs[[5]] <- current$edit.mtx
add.edge(3, 1)
graphs[[6]] <- current$edit.mtx
add.edge(6, 4)
graphs[[7]] <- current$edit.mtx
add.edge(2, 5)
graphs[[8]] <- current$edit.mtx
add.edge(2, 3)
graphs[[9]] <- current$edit.mtx
add.edge(4, 3)
graphs[[10]] <- current$edit.mtx
add.edge(6, 1)
graphs[[11]] <- current$edit.mtx


fn <- '../figures/non_expert_empty'
titles <- plot.cv(graphs, data, k, fn)

plot.r2(titles) + ylim(0, 1)
ggsave(paste0(fn, '_R2.pdf'), device='pdf', height=4, width=5)
```

